Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-05 paper/paper.tex
Edits for Section 5

In this section, we present the dependent type theory \lang{}, whose logical
consistency will be proven in Section~\ref{sec:logrelproof}.

The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
type theory, terms and types are collapsed into the same syntactic
category. We use a Curry-style (extrinsic) syntax, where terms are not
annotated by types, and only terms which participate in reduction are part of
introduction and elimination forms.

@@

\begin{figure}
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Set, Void, Bool, Eq, Red}
\[ \drule[width=5in]{I-Pi} \]
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}

The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation in Figure~\ref{fig:logrel}.
% The logical relation takes the form $[[Interp I i A S]]$.
The metavariables $[[A]]$ and $[[i]]$ are terms and naturals, respectively.
% \scw{Many introductory texts define the relation as a recursive function
% over type structure, or step-indices. You use an inductive relation instead, why?}
%\yl{resolved}
% \scw{Can we view this inductive relation as the graph of the partial function
% that is defined recursively over types?} \yl{reolsved}
% \scw{What is this form extensible too? impredicative quantification? recursive
% types? } \yl{resolved? impredicativity is hard}
% \scw{Is it worth observing here that this definition is not over sets of typed
% terms. That it characterizes all terms that look like booleans (i.e. evaluate to
% true or false) or all terms that look like proofs (i.e. evaluate to refl). The
% fact that there is no connection between p and a and b in the I-Eq case is strange
% looking. Need to explain.  }
% \yl{resolved}
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures,
using $[[PowerSet STm]]$ to denote the powerset of the set of terms.
\[
    [[I]] \in [[ { j | j < i  } ->  PowerSet STm ]] \qquad\qquad\qquad
    [[S]] \in [[PowerSet STm]]
\]
The function $[[I]]$ is a family of sets of terms indexed by natural numbers
strictly less than $[[i]]$, which represents the current universe level.
In \rref{I-Set}, the function $[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} ensures predicativity of the system.

@@

\section{Existence of \texorpdfstring{$\beta$}{beta}-normal forms}
\label{sec:extension}
In this section, we show how the logical relation from
Section~\ref{sec:logreldep} can be extended to show the existence of $\beta$
normal forms for (open and closed) well-typed terms.  In other words, we prove
that it is possible to repeatedly use the parallel reduction relation to
reduce any term to its unique normal form, where no further (non-identity)
reductions can be applied. This result can be used to show that our type
conversion relation is decidable.% , as computation of the normal form will
% always terminate.\yl{dependent on reduction strategy}

The goal of this section is also to demonstrate that our logical relations
proof technique can be extended to reason about the reduction properties of
open terms, not just the reduction of terms after closing substitutions.
Reasoning about open terms is particularly important for dependently-typed
languages because type checking involves working with open terms.
\scw{
  Add when we can find a reference:
  However, even non dependently-typed languages employ such techniques,
  especially in the case of relational semantics.
}
While this extension employs well-known techniques, it continues to be short and
demonstrates the robustness of our initial framework.

We begin this part with a description of the $\beta$-normal forms
of \lang{}.
%
\begin{figure}
  \[
    \begin{array}{llcl}
       \beta\text{-}\mathit{neutral\ terms} &
      [[e]] & ::= & [[x]]\ |\ [[e f]]\ |\ [[J e f]]\ |\ [[if e f
                    f]] \\ \\
      \beta\text{-}\mathit{normal\ terms} &
      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi x : f . f]]\
                    |\ [[f ~ f : f]]\\
            & & |   & [[\ x . f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
    \end{array}
  \]
  \caption{$\beta$-neutral and normal forms}
  \label{fig:nenf}
\end{figure}
%
The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the
neutral terms and normal forms with respect to $\beta$-reduction\scw{Why not say parallel reduction here? We can be specific and say that that the only reductions available for these terms are identity reductions and cite
\footnote{\dotv{normalform.v}{nf\_refl}} }.\yl{I always think of
normal form as the specific definition that says the relation can't
step. We can refer to nfrefl but that requires some explanation about
the definition of normal form and the complication that it doesn't
hold in $\eta$ (maybe it's fine if we just don't mention it in the
$\eta$ case)}\scw{A terminal form is a syntactic characterization of terms that don't step 
according to a particular relation. A normal form is a terminal form of a normalizing relation.}
 Instead of the
metavariables $e$ and $f$, we also
use the judgment forms $[[ne a]]$ and $[[nf a]]$ to indicate that there exists
$[[e]]$ or $[[f]]$ such that $[[a = e]]$ or $[[a = f]]$.

The predicates $[[wne a]]$ and $[[wn
a]]$ describe terms that can evaluate into $\beta$-neutral or
$\beta$-normal form through parallel reduction and are defined as
follows.
\[
\begin{array}{llcl}
\mbox{\emph{weakly normalizes to a neutral form}} & [[wne a]] &\iff& \exists [[e]], [[a =>+ e]] \\
\mbox{\emph{weakly normalizes to a normal form}}  & [[wn a]]  &\iff& \exists [[f]], [[a =>+ f]] \\
\end{array}
\]

\begin{figure}[H]
  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
  \caption{Extended logical relation (new and changed rules) }
  \label{fig:logrelopen}
\end{figure}

The updated logical relation is shown in
Figure~\ref{fig:logrelopen}.~\footnote{\dotv{semtypingopen.v}{InterpExt}}
There is one new rule in this figure, \rref{I-Ne}. In a non-empty context, a
type itself may evaluate to a neutral term and in turn can only be inhabited
by neutral terms.  Otherwise, the rest of the rules in this figure are updates
to the analogous rules in Figure~\ref{fig:logrel}. Note that, we omit the
rules for the function and universe cases because they are identical to the
original version.

% The idea of a term blocked from $\beta$ reduction is
% captured by $[[e]]$, the set of neutral terms. We augment the
% interpretation of booleans and the empty types with terms that
% evaluate to neutral terms. Furthermore, types that are neutral terms
% can also be assigned a meaning since they may be inhabited by other
% neutral terms.
The changes to \rref{I-Bool} and \rref{I-Void}
follow the same pattern: an open term of type $[[Bool]]$ does not necessarily
reduce to $[[true]]$ or $[[false]]$, but may reduce to a variable, or more
generally, a neutral term. Likewise, while the $[[Void]]$ type remains
uninhabited under an empty context, it may be inhabited when there is a
variable in the context that has type $[[Void]]$ or that can be eliminated to
type $[[Void]]$.

\scw{Cassia didn't understand this part}
The rule for equality type $[[a ~ b : A]]$ is augmented with the precondition that
$[[a]]$, $[[b]]$, and $[[A]]$ are all normal forms because
otherwise our model would include equality types that are themselves
not normalizing.
Furthermore, the condition $[[a <=> b]]$ is only
required when the equality proof reduces to $[[refl]]$. If the proof
term reduces to a neutral term, then there is nothing we need to show
about the relationship between $[[a]]$ and $[[b]]$.

>>>>

In this section, we present the dependent type theory \lang{}, whose logical
consistency will be proven in Section~\ref{sec:logrelproof}.
Its syntax is given in Figure~\ref{fig:syntax}. As a dependent
type theory, terms and types are collapsed into the same syntactic
category. We use a Curry-style (extrinsic) syntax, where terms are not
annotated by types, and only terms which participate in reduction are part of
introduction and elimination forms.

@@

\begin{figure}
\drules[I]{$[[Interp I i A S]]$}{Logical relation}{Set, Void, Bool, Eq, Red}
\[ \drule[width=5in]{I-Pi} \]
\caption{Logical relation for \lang{}}
\label{fig:logrel}
\end{figure}

The logical relation\footnote{\dotv{semtyping.v}{InterpExt}} for \lang{},
which takes the form $[[Interp I i A S]]$,
is defined as an inductively generated relation in Figure~\ref{fig:logrel}.
% The logical relation takes the form $[[Interp I i A S]]$.
The metavariables $[[A]]$ and $[[i]]$ are terms and naturals, respectively.
% \scw{Many introductory texts define the relation as a recursive function
% over type structure, or step-indices. You use an inductive relation instead, why?}
%\yl{resolved}
% \scw{Can we view this inductive relation as the graph of the partial function
% that is defined recursively over types?} \yl{reolsved}
% \scw{What is this form extensible too? impredicative quantification? recursive
% types? } \yl{resolved? impredicativity is hard}
% \scw{Is it worth observing here that this definition is not over sets of typed
% terms. That it characterizes all terms that look like booleans (i.e. evaluate to
% true or false) or all terms that look like proofs (i.e. evaluate to refl). The
% fact that there is no connection between p and a and b in the I-Eq case is strange
% looking. Need to explain.  }
% \yl{resolved}
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures,
using $[[PowerSet STm]]$ to denote the powerset of the set of terms.
\[
    [[I]] \in [[ { j | j < i  } ->  PowerSet STm ]] \qquad\qquad\qquad
    [[S]] \in [[PowerSet STm]]
\]
The function $[[I]]$ is a family of sets of terms indexed by natural numbers
strictly less than $[[i]]$, which represents the current universe level.
In \rref{I-Set}, the function $[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} ensures predicativity of the system.

@@

\section{Existence of \texorpdfstring{$\beta$}{beta}-normal forms}
\label{sec:extension}

In this section, we extend the logical relation in Figure~\ref{fig:logrel}
to show the existence of $\beta$-normal forms for both open and closed
well-typed terms. More precisely, we prove it possible to repeatedly apply
parallel reduction to reduce a term to its unique normal form. Consequently,
this shows that the convertibility relation is decidable.

This extension of the logical relation demonstrates that our proof technique
can also be used to reason about the reduction properties of open terms, not
just of terms after closing substitutions. Reasoning about open terms is
especially important for dependently typed languages, as type checking involves
working with open terms.
\scw{
  Add when we can find a reference:
  However, even non dependently-typed languages employ such techniques,
  especially in the case of relational semantics.
}
While the extension use well-known techniques,
\jc{Does this need a citation?}
it remains short and demonstrates the robustness of our initial framework.

\begin{figure}
  \[
    \begin{array}{llcl}
       \beta\text{-}\mathit{neutral\ terms} &
      [[e]] & ::= & [[x]]\ |\ [[e f]]\ |\ [[J e f]]\ |\ [[if e f
                    f]] \\ \\
      \beta\text{-}\mathit{normal\ terms} &
      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi x : f . f]]\
                    |\ [[f ~ f : f]]\\
            & & |   & [[\ x . f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
    \end{array}
  \]
  \caption{$\beta$-neutral and normal forms}
  \label{fig:nenf}
\end{figure}

Figure~\ref{fig:nenf} gives the $\beta$-neutral forms $[[e]]$ and
$\beta$-normal forms $[[f]]$ of \lang{}. Neutral terms consist of variables and
elimination forms that do not reduce any further, while normal terms consist of
neutral terms and other introduction forms. We use the judgment forms $[[ne a]]$
and $[[nf a]]$ to indicate that there exists a term $[[e]]$ (resp. $[[f]]$)
such that $[[a = e]]$ (resp. $[[a = f]]$). Additionally, we define predicates
to describe terms that weak normalization: reduction to neutral or normal forms
by parallel reduction.

% The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the
% neutral terms and normal forms with respect to $\beta$-reduction\scw{Why not say parallel reduction here? We can be specific and say that that the only reductions available for these terms are identity reductions and cite
% \footnote{\dotv{normalform.v}{nf\_refl}} }.\yl{I always think of
% normal form as the specific definition that says the relation can't
% step. We can refer to nfrefl but that requires some explanation about
% the definition of normal form and the complication that it doesn't
% hold in $\eta$ (maybe it's fine if we just don't mention it in the
% $\eta$ case)}\scw{A terminal form is a syntactic characterization of terms that don't step 
% according to a particular relation. A normal form is a terminal form of a normalizing relation.}

\begin{definition}[Weak normalization]
  We define weak normalization of a term $[[a]]$ to a neutral form, written
  $[[wne a]]$, or to a normal form, written $[[wn a]]$, as the following:
  \begin{align*}
    [[wne a]] &\iff \exists [[e]], [[a =>+ e]] \\
    [[wn a]]  &\iff \exists [[f]], [[a =>+ f]]
  \end{align*}
\end{definition}

\begin{figure}
  \drules[I]{$[[Interp I i A S]]$}{Extended Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
  \caption{Extended logical relation for \lang{} (new and changed rules)}
  \label{fig:logrelopen}
\end{figure}

The modifications to the logical relation\footnote{\dotv{semtypingopen.v}{InterpExt}}
are given in Figure~\ref{fig:logrelopen}. The new \rref{I-Ne} asserts that
only terms that reduce to neutral forms inhabit the interpretation of neutral
types. The remaining rules are updates to existing rules; we omit those for
function types and universes as they are unchanged.

The changes in \rref{I-BoolNew} and \rref{I-VoidNew} follow the same pattern. An open
term of type $[[Bool]]$ may reduce to $[[true]]$ or $[[false]]$ as before, but
may also reduce to a neutral term, such as a variable. Likewise, while in an
empty context the interpretation of $[[Void]]$ remains empty, if the context
is not empty, then it may contain, for instance, a variable in the context of
type $[[Void]]$ or that can be eliminated into type $[[Void]]$.

In \rref{I-EqNew}, we add preconditions asserting that $[[a]]$, $[[b]]$, and
$[[A]]$ are normal so that we only model normalizing equality types.
Furthermore, we only require $[[a <=> b]]$ when the equality proof reduces to
$[[refl]]$. If the proof term reduces to a neutral term, there is nothing we
can assert about the relationship between $[[a]]$ and $[[b]]$.
